---
title: "Introduction to Parallel Computation in R"
author: "Brenton Kenkel"
date: "May 6, 2015"
thanks: >
  This document and all associated content are licensed under the [Creative Commons Attribution-ShareAlike License (v4.0)](http://creativecommons.org/licenses/by-sa/4.0/).  Source code is available at <https://github.com/brentonk/parallel-r-workshop>.  The document incorporates material from [Jonathan Olmsted's Advanced Statistical Programming Camp notes](http://olmjo.com/teaching/ASPC/).
output:
  pdf_document:
    highlight: tango
    template: template.tex
    keep_tex: true
    fig_caption: true
---

# Introduction

# Running Example

As a running example throughout the workshop, we'll be replicating part of the analysis in Josh Clinton's article ["Representation in Congress: Constituents and Roll Calls in the 106th House"](http://www.jstor.org/stable/10.1111/j.1468-2508.2006.00415.x) (*Journal of Politics*, 2006).  The paper is concerned with whether (and how) the preferences of the constituents in a Congress member's district affect her vote choices.  The part of the analysis we'll be replicating is summarized in Table 4 of the paper.  The analysis consists of running more than 800 logistic regression models, each with a different roll call vote as the dependent variable, on three covariates: an indicator for the member's party ID, the average ideology of the member's same-party constituents, and the average ideology of the member's different-party constituents.  The goal of the analysis is to track the number of models in which each covariate is statistically significant.

![Table 4 from Clinton 2006, summarizing the analysis we'll be partially replicating.](clinton-table-4.pdf)

You can download a cleaned-up version of Josh's data from my website in R data format:

```{r load}
load(url("http://bkenkel.com/data/clinton-jop.rda"))
```

The data file contains two objects, `house106` and `votes106`.  `house106` is a $`r nrow(house106)` \times `r ncol(house106)`$ data frame containing legislator characteristics—the independent variables in each of our logistic regression models.

```{r describe-house}
str(house106)
```

`votes106` is a $`r nrow(votes106)` \times `r ncol(votes106)`$ roll call matrix, where the $ij$'th entry gives legislator $i$'s vote on bill $j$.[^nvotes]

[^nvotes]: To avoid being derailed by spurious warnings during the workshop, I removed votes for which there was separation in the corresponding logistic regression, so the number of votes analyzed is slightly smaller than in the original paper.

```{r describe-votes}
str(votes106)
```

So we're going to run `r ncol(votes106)` logits: one for each column of `votes106`, each with the three variables in `house106` as regressors.  As an example, let's do that with the first vote recorded in `votes106`.

```{r first-logit}
## Substitute given vote into member data
house106$vote <- votes106[, 1]

## Run logit of vote choice on member attributes
logit_fit <- glm(vote ~ republican + prefs_same + prefs_nonsame,
                 data = house106,
                 family = binomial(link = "logit"))
```

# Preliminaries

```{r packages, eval=FALSE}
install.packages(c("microbenchmark", "foreach", "doParallel"))
```

## Functions

```{r function}
pval_by_vote <- function(vote, coefficient) {
    ## Substitute given vote into member data
    house106$vote <- votes106[, vote]

    ## Run logit of vote choice on member attributes
    logit_fit <- glm(vote ~ republican + prefs_same + prefs_nonsame,
                     data = house106,
                     family = binomial(link = "logit"))

    ## Retrieve p-value
    p_vals <- summary(logit_fit)$coef[, 4]
    return(p_vals[coefficient])
}
```

```{r function-test}
pval_by_vote(vote = 10, coefficient = 2)
```

## Loops

```{r for}
p <- rep(NA, 5)
for (i in 1:5) {
    p[i] <- pval_by_vote(vote = i, coefficient = 2)
}
```

```{r for-results}
sum(p < 0.05)
```

```{r for-function}
pvals_for <- function(votes, coefficient) {
    p <- rep(NA, length(votes))
    for (i in 1:length(votes)) {
        p[i] <- pval_by_vote(vote = votes[i], coefficient = coefficient)
    }
    return(p)
}
```

```{r for-function-results}
sum(pvals_for(1:5, 2) < 0.05)
```

## Benchmarking

```{r system-time-applied}
system.time(pvals_for(1:5, 2))
```

```{r microbenchmark, cache=TRUE}
library("microbenchmark")
microbenchmark(five = pvals_for(1:5, 2),
               ten = pvals_for(1:10, 2))
```

# From `for` to `foreach`

```{r foreach}
library("foreach")
```

```{r foreach-syntax}
foreach (i = 1:5) %do% {
    pval_by_vote(vote = i, coefficient = 2)
}
```

```{r foreach-combine}
foreach (i = 1:5, .combine = "c") %do% {
    pval_by_vote(vote = i, coefficient = 2)
}
```

```{r foreach-function}
pvals_do <- function(votes, coefficient) {
    foreach (i = votes, .combine = "c") %do% {
        pval_by_vote(vote = i, coefficient = coefficient)
    }
}
```

```{r for-vs-do, cache=TRUE}
microbenchmark(for_loop = pvals_for(1:5, 2),
               do = pvals_do(1:5, 2))
```


# Local Parallelization

```{r dopar-syntax}
foreach (i = 1:5, .combine = "c") %dopar% {
    pval_by_vote(vote = i, coefficient = 2)
}
```

```{r dopar-function}
pvals_dopar <- function(votes, coefficient) {
    foreach (i = votes, .combine = "c") %dopar% {
        pval_by_vote(vote = i, coefficient = coefficient)
    }
}
```

```{r dopar-setup}
library("doParallel")
cl <- makePSOCKcluster(2)
registerDoParallel(cl)
getDoParWorkers()
```

```{r cluster-export}
clusterExport(cl, varlist = ls())
```

```{r small-test, cache=TRUE}
microbenchmark(for_loop = pvals_for(1:5, 2),
               do = pvals_do(1:5, 2),
               dopar = pvals_dopar(1:5, 2))
```

```{r medium-test, cache=TRUE}
cols <- 1:ncol(votes106)
system.time(p_do <- pvals_do(cols, 2))
system.time(p_dopar <- pvals_dopar(cols, 2))
```

```{r check-equality}
all.equal(p_do, p_dopar)
```

```{r get-answer}
sum(p_dopar < 0.05)
```

```{r big-test, cache=TRUE}
system.time(for (j in 1:4) pvals_for(cols, j))
system.time(foreach (j = 1:4) %dopar% pvals_for(cols, j))
```

```{r stop-cluster}
stopCluster(cl)
```

## For Mac and Linux Users: An Easier Way


# Remote Parallelization
